How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?

### 2021-04-01 analysis/paper/paper.Rmd
Editing the paper

---
title: "How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?"
author:
  - Gayoung Park
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2: 
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Human behavioural ecological models use stone artefacts as a source of information for understanding forager mobility. We use these models to examine stone artefacts from 22 sites in Korea to investigate mobility and site occupation during the Late Pleistocene. This study focuses on two questions: how did the introduction of stemmed points change forager' mobility and landscape use? And how did this new technology change the way people use habitation sites? We analyse artefact volumetric density, retouch frequency, composition of toolkits, artefact raw materials, environmental and demographic contexts. Quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages before and after the introduction of stemmed points throughout the Late Pleistocene. Our results show that forager groups using stemmed points were associated with occupation of marginal or extreme environments. In addition, groups with stemmed points were associated with expedient technologies indicating residential and less mobile behaviours. The environmental context of this technological innovation was a gradual decrease in temperature into the LGM. A population increase followed after the appearance of stemmed points.

@@

The appearance of stemmed points (Figure \@ref(fig:stemmed-points)) in the Late Pleistocene (\~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it reflected more specialized hunting practices [@seong2008tanged; @chang2013human]. Stemmed points are often called as tanged points in the western hemisphere, but we use 'stemmed point' to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea. The stemmed point is a projectile point made out of elongated flake or blade with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. Stemmed points are likely to have been projectile weapon tips because the stem of these tools is considered to represent hafting technology that joins the lithic to a wooden shaft. If they were projectile weapon tips, stemmed points represent the first appearance of hafting technology in Korea. Most Korean Paleolithic sites do not preserve organic remains, so we lack direct fauna evidence on hunting and subsistence strategies. Here we explore questions about the implications of the appearance of this new technology using a behavioural ecological framework and stone artefacts. We explore changes in the way people occupied the landscape, the way people used sites, and impact of social and ecological environment. We examined stone artefacts from 22 sites in South Korea to investigate the changes in human behaviour associated with the appearance stemmed points, and the blade industry that followed.

@@

Previous work on this technological change in Korea has focussed on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin]. To our knowledge, the implications of the appearance of stemmed points for forager mobility and site use have not been explored in detail. Here, we attempt to identify patterns of mobility through analysis of the stone artefact assemblages using Human Behavioural Ecology (HBE) models. Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018mobility], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.

@@

In the Eastern hemisphere, major technological innovations during the Late Pleistocene are restricted to Northeast Asia, such as northern China, Japan and Korea. These innovations include blade and micro technology, high frequencies of retouched blade tools, several novel tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary]. Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals [@seong2008tanged; @bae2010origin; @bae2017origin].

The Shuidonggou site in northern China is an important example of the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou]. Shuidonggou Locality SDG 2, which dates to around c. 32,000 years, contains fauna specimens, ostrich eggshell beads, and stone artefacts including blades, retouched flakes, cores, and debris along with well-preserved hearths. In Locality 9 dated to c. 29,000 years, there is a small scatter of stone artefacts consisting of blades, Levallois flakes, cores and other retouched flakes. The recent excavation of Locality 12 shows that its archaeological context, dated to 13,078--13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks. In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [@pei2012shuidonggou; @gao2014discovery; @zhang2016bone]. Stemmed points are not part of the Shuidonggou assemblages, but microblade technology is argued to have appeared in China after 29ka BP, related to changes in mobility associated with colder climates towards LGM [@yi2014significance]. Shuidonggou shows that the blade component came from Siberia and/or Mongolia at 41 ka (SDG Locality1 and 9), advanced core and flake tools were likely to have been locally developed at 33 ka (2 and 8), and at 10.8 ka microblade technology appears but the origin is uncertain (12) [@yi2014significance; @li2019history].

Late Pleistocene technological innovations in Japan appeared in different regions of the archipelago around 38,000 years ago accompanied by remains of Homo sapiens found in Okinawa Island (c.36,000 cal BP) [@yamaoka2012use; @izuho201421]. These new technologies include trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak-shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [@bae2017origin; @izuho201421]. These innovations have been interpreted as evidence of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, driven by increased population and ecological changes [@yamaoka2012use; @bae2017origin; @nakazawa2018quaternary; @morisaki2019early]. Hafted trapezoids were likely multifunctional tools adapted to the specific environmental settings of the different Japanese Islands [@ono2002radiocaron; @yamaoka2012use]. Ishinomoto 8-ku is one of the earliest sites dated to 39,690--34,790 cal BP, located in Takuma upland of Kumamoto Prefecture. A total of 500 stone artefacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools. The main lithic raw materials are chert and andesite [@izuho201421]. Stemmed points appear much later, dated to between about 15,500 and 13,800 cal BP [@ono2002radiocaron; @tsutsumi2007dynamics].

What makes the Korean Late Pleistocene technological transition distinctive from what we see at Shuidonggou and in Japan is that the earliest signs of new technologies in Korean assemblages are stemmed points, followed by blade technologies. Stemmed points appeared in both regions after the case of Korean Peninsula. The oldest stemmed points so far is from the Yonghodong site dated to 38.5 ka and made on elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin]. After blades appeared at c. 27 ka and were adopted to the stemmed points, the shape of points became more standardized with one or two ridges on the dorsal side and triangular cross section. Replacing flake to blade as a blank of the points led to an increase in quantity over time. In addition to the stemmed points, flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene [@bae2010origin; @lee2013current; @Lee_2016]. There are several attempts to explain the technological transition and they can be summarized into two competing models: 'heterogenic' migration [@bae2010origin] or *in situ* evolution [@seong2009emergence]. The migration model argues that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins as the result of continuing influx of model human migration from two routes. Bae (2010) explains that blade and stemmed points arrived in the Korean peninsula from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP with the earliest evidence of blade in those regions. He assumes that core and flake industry come from southern China based on the existence of similar assemblages in both Korea and southern China in addition to a genetic analysis of the Y-chromosome of the modern population [@bae2013early; @lee2013current]. The '*in situ*' model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the south of the Korean peninsula with no external influence [@seong2008tanged]. This claim is supported by the earliest appearance of stemmed points among Northeast Asia, such as Hwadae-ri, Hopyeong-dong, and Yonghodong sites dated to 40-35ka BP [@seong2009emergence]. Seong(2009) uses the increased blade-to-flake ratios on lithic assemblages to support the *in situ* model based on the premise that the blade industry represents new technology while flakes indicate a continuing local one.\

<!-- need to put two images,  one for stemmed points, one for overall lithic assemblages -->

@@

Our brief review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focussed on timing, location, origins and description of these new tools, but modelling the land-use behaviours that generated these assemblages has rarely been undertaken. We focus on the appearance of stemmed points in Korea because these are a new technology that represents a new hunting method. This is because the hafting implies throwing techniques that can reach long-distance targets [@Keeley_1982; @kuhn2015artifacts]. We draw on behavioural ecological theory to model the effects of this technological change. Behavioural ecology theory offers structured and testable models based on optimality assumptions that stand on economic rationality and environmental knowledge [@winterhalder1992evolutionary; @prentiss2019handbook]. The patch choice model and marginal value theorem predict that forager mobility and travel times will increase when resource patch productivity decreases [@smith1991inujjuamiunt; @smith1983anthropological; @bettinger1997evolutionary; @wolverton2015patch; @llano2015optimal]. We can also model individual artefact types as a kind of resource patch. Foragers may find a more costly and complicated technology (such as stemmed points relative to flakes) optimal if they have a long cumulative time in use, if they are maintainable and reusable in a landscape where lithic raw material supply is uncertain. That is, they may move to a patch where they can stay longer (i.e. an artefact type that has a long and extendible use-life) when they are not certain of the productivity and travel times of other patches [@kuhn2015artifacts].

We ask two questions to investigate the period around the introduction of stemmed points in Korea. First, how did the introduction of the new tools change foragers' landscape use? Specifically, was the use of stemmed points associated with occupation of marginal or extreme environments? Our assumption here is that this new technology might have allowed foragers to explore and sojourn in less productive landscapes by increasing their mobility. In Australia during the mid-Holocene, Hiscock et al. argue that the first appearance of microlithics in many locations represent a portable and multi-functional toolkit that minimize travel expenses and increase tool readiness to adapt to patchy and unpredictable resources associated with environmental change [@Hiscock_1994; @Hiscock_Clarkson_Mackay_2011].Similarly, in Japan, Morisaki et al. [@Morisaki_Izuho_Terry_Sato_2015] argue that the transition from trapezoid to the blades and projectile points around 25 ka cal BP in north Paleo-Honshu Island enabled the foragers to extend their occupation into cold grassland landscapes.

Second, we ask how the new technology changed the way people use habitation sites and mobility. After the appearance of these new tools, did people tend to stay in one location for longer or shorter periods, perhaps for specific purposes? The stone artefact assemblage from a site can inform us how long individuals or groups stayed and give insights into their activities [@binford1979organization; @kelly1995foraging; @kuhn2016moves; @holdaway2019surface]. For example, stone artifact density and the frequency of retouched pieces (scaled to the volume of excavated sediment) can be used as proxies to represent occupation patterns [@barton2011modeling; @clark2017lithics]. Assemblages with low density but a high proportion of retouched and backed pieces may indicate the remains of 'short-term camp', which is a site for small and ephemeral overnight camp or limited activity station. On the other hand, the combination of high density and low retouched pieces represent logistically organized basecamp, which is a site with greater residential stability, long site occupation and occupied by larger groups [@clark2017lithics; @clark2019landscapes].

@@

After the first excavation of a Paleolithic site in Korea at Seokjang-ri site in the 1960s, more than 200 Paleolithic sites have been discovered in South Korea [@lee2019were]. We selected sites dated to before and after the transition period (40-35 ka) to analyze the process of technological change. This resulted in a sample of analysed 33 assemblages from 22 sites spanning 49-24 ka (Figure \@ref(fig:korea-map)). We identified multiple assemblages in a site where culturally sterile deposits separated artefact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artefacts. For example, the Hwadae-ri site has three cultural layers. The lowest horizon at the bottom, dated to 39,000±1400 BP by OSL, contains coarse flake tools made of vein quartz and quartzite. The middle cultural horizon has stemmed points made of porphyry, with quartz and quartzite dominating the assemblage. This layer was dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL. The uppermost layer, dated to 22,000±100 BP by OSL, contains blades, scrapers, awls, and denticulates [@seong2009emergence]. We obtained the data from published excavation reports held by provincial museums.

@@

The 33 assemblages consist of cores, blade cores, blades, flakes, debris, hammers, choppers, plans, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc. We excluded some artefacts from our sample because of uncertainty about their typology. The stone artefacts from all of these sites are stored in museums located throughout South Korea, and have been briefly described in the excavation reports. We have visited most of these collections to obtain permission to study the materials.

@@

The artefact volumetric density is defined as the total number of artefacts per cubic meter of excavated sediment, and serves as a proxy for the accumulation ratio of artefacts [@clark2017lithics; @clark2019landscapes]. We used these density values to evaluate mobility and land-use practices of hunter-gatherers during the Late Pleistocene. The retouch frequency is the number of retouched tools divided by the total number of artefacts in an assemblage. This value represents mobility and duration of site occupation based on the concept that tool reduction is an tactic to extend the use-life of a tool by generating sharp, usable edges while minimizing the cost of transporting tools or bulky raw materials by reducing the total load weight [@Kuhn_1990; @buvit2014late; @clark2019landscapes]. In other words, producing and carrying highly retouched tools will increase portability and efficiency by decreasing the size of both individual tools and the assemblages so that hunter-gatherers could move often or further, or both [@kuhn1991unpacking; @kuhn1994formal; @kuhn2004upper; @andrefsky1994raw; @davies2018modeling]. Combining these two proxies, the artefact volumetric density and the retouch frequency, we expect to determine whether or not the hunter-gatherers were residentially mobile or logistically organized. If the assemblage in a site has a high artefact density and small proportion of retouch, we interpret it as a more expedient assemblage that represents "base camps" or "residences", while less density with higher proportions of retouch shows a more curated assemblage and indicates either residential mobility or certain task groups away from residential sites [@riel2004late; @clark2017lithics].

@@

Studying the raw material consumption patterns is based on an assumption that stone raw materials were transported and consumed to optimize mobility and minimize risk of not having artefacts when they were needed [@andrefsky1994raw; @brantingham2000raw; @brantingham2003neutral; @yue2020lithic]. The representation of raw materials in an assemblage can be an indicator of site occupation patterns. For example, a large variability of raw materials in a site can suggest long-term settlement, while varied and exotic raw materials can indicate higher mobility [@valde2020models]. Here we want to know the specific raw materials that relate to the appearance of a new technology, and understanding site occupation patterns by examining changes in the proportions of raw material types in the assemblages.

HBE models require environmental proxies to build hypotheses of behavioural change influenced by the surrounding environment. However, there are few paleoenvironmental proxies in Korean relevant to our target period. For example, geoarchaeological observations suggest that environment of the period was "overall cooler and drier during MIS 3" [@Han_2008; @seong2008tanged; @Choi_2011; @chang2013human; @im2015study; @bak2017late]. To supplement the limited availability of high-resolution local proxies, we used climate information from the simulated data set of paleoclimate including global monthly temperature, covering the last 120,000 years [@beyer2020high]. We applied information of site location including elevation, latitude and longitude to the simulated data set to get the annual temperature of our research period and area. Besides the climate analysis, we analysed site elevation as a proxy to represent different environments at local scales, to identify how these impact mobility and site occupation. We mainly focused on the elevation of the sites before and after the appearance of new technology and its relationship with the stemmed points.

We used radiocarbon dates as a reference for population level. Summed probability distributions (SPD) of radiocarbon dates are often used to understand temporal trends in past human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed; @riris2019widespread]. Despite limitations of the validity of SPD [@williams2012use; @bamforth2012radiocarbon], including the lack of excavated sites in certain periods, sample size, and calibration effects, this method may be helpful as a proxy for a first approximation of population fluctuations. Although this is often used with samples of thousands of ages, this method can also be useful for small sample sizes. For instance, Timpson et al. apply SPD to 93 ages from Eastern Middle Sweden and demonstrate that the result is equivalent to a larger sample size (n= 243) [@timpson2014reconstructing]. We used 108 radiocarbon ages from 22 sites to examine connections between population trends during the Late Pleistocene in Korea and technological innovation.

>>>>

---
title: "How did the introduction of Stemmed Points affect Mobility and Site Occupation during the Late Pleistocene in Korea?"
author:
  - Gayoung Park
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2: 
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Human behavioral ecological models, under the umbrella of cultural evolution, use stone artefacts as a source of information for understanding forager subsistence strategies. We use these models to examine stone artefacts from 22 sites in Korea to investigate mobility and site occupation patterns during the Late Pleistocene. This study focuses on two questions: what changes in foragers' landscape use were associated with the introduction of the new tools? And what changes in the way people use habitation sites and mobility were associated with the new technology? We conduct quantitative analyses including artefact volumetric density, retouch frequency, composition of toolkits, and artefact raw materials, and explore environmental and demographic contexts by applying paleoclimate simulation and summed probability distribution models. Quartz and side scrapers, in addition to cores and choppers, remain dominant in assemblages before and after the introduction of stemmed points throughout the Late Pleistocene. Our results show that forager groups using stemmed points were associated with occupation of marginal or extreme environments. In addition, groups with stemmed points were associated with expedient technologies indicating residential and less mobile behaviours. The environmental context of this technological innovation was a gradual decrease in temperature into the LGM. A population increase followed after the appearance of stemmed points.

@@

The appearance of stemmed points (Figure \@ref(fig:stemmed-points)) in the Late Pleistocene (\~40-35 ka) in Korea is often assumed to have transformed forager lifeways because it reflected more specialized hunting practices [@seong2008tanged; @chang2013human]. Stemmed points are often called tanged points in the western hemisphere, but we use 'stemmed point' to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea. The stemmed point is a projectile point made from an elongated flake or blade with slight retouch on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a shaft. Stemmed points are likely to have been projectile weapon tips because the stem of these tools is considered to represent hafting technology that joins the lithic to a wooden shaft. If they were projectile weapon tips, stemmed points represent the first appearance of hafting technology in Korea. Most Korean Paleolithic sites do not preserve organic remains, so we lack direct fauna evidence on hunting and subsistence strategies. Here we explore questions about the implications of the appearance of this new technology using a behavioural ecological framework and stone artefacts. We explore changes in the way people occupied the landscape, the way people used sites, and impact of social and ecological environment. We examined stone artefacts from 22 sites in South Korea to investigate the changes in human behaviour associated with the appearance stemmed points, and the blade industry that followed.

@@

Previous work on this technological change in Korea has focused on issues of the transition from the Early to the Late Paleolithic and modern human dispersals [@norton2009evolution; @bae2010origin]. To our knowledge, the implications of the appearance of stemmed points for forager mobility and site use have not been explored in detail. Here, we attempt to identify patterns of mobility and site occupation through analysis of the stone artefact assemblages using Human Behavioural Ecology (HBE) concepts. Drawing on established relationships between lithic technology and forager mobility [@shott1986settlement; @Hiscock_1994; @kuhn1994formal; @kuhn2004upper; @kuhn2015artifacts; @kuhn2016moves; @capriles2018mobility], we hypothesize that the appearance of the new hunting tool might reflect a preference for more portable and efficient technologies, that are part of a broader strategy of moving frequently and further, possibly as an adaptation to environmental or population changes.

@@

In the Eastern hemisphere, major technological innovations during the Late Pleistocene are restricted to Northeast Asia, such as northern China, Japan and Korea. These innovations include blade technology, high frequencies of retouched blade tools, several novel tools such as projectile points, stemmed points, end scrapers, burins, denticulates, etc, and using high-quality raw materials [@bar1999big; @brantingham2001initial; @Bar-Yosef_2002; @bae2017origin; @nakazawa2018quaternary]. Previous work has argued that the appearance of new stone artefact technologies in this region may be linked to modern human dispersals [@seong2008tanged; @bae2010origin; @bae2017origin].

The Shuidonggou site in northern China is an important example of the transition from cobble to blade tool industries in the Late Pleistocene of Northeast Asia [@Brantingham_Perreault_2010; @Gao_Zhang_Yang_Shen_Wu_2010; @pei2012shuidonggou]. Shuidonggou Locality SDG 2, which dates to around c. 32,000 years, contains fauna specimens, ostrich eggshell beads, and stone artefacts including blades, retouched flakes, cores, and debris along with well-preserved hearths. In Locality 9 dated to c. 29,000 years, there is a small scatter of stone artefacts consisting of blades, Levallois flakes, cores and other retouched flakes. The recent excavation of Locality 12 shows that its archaeological context, dated to 13,078--13,296 cal BP, includes ground stone and more than 30,000 microlithics made out of a variety of raw materials such as fine-grained and siliceous rocks. In addition, more than 10,000 animal fossils and bone tools including a tool for fishing nets, an awl, and two needles were excavated [@pei2012shuidonggou; @gao2014discovery; @zhang2016bone]. Shuidonggou shows that the blade component came from Siberia and/or Mongolia at 41 ka (SDG Locality1 and 9), advanced core and flake tools were likely to have been locally developed at 33 ka (2 and 8), and at 10.8 ka microblade technology appears but the origin is uncertain (12) [@yi2014significance; @li2019history]. The microblade technology is argued to have appeared related to changes in mobility associated with colder climates towards LGM [@yi2014significance].

The Japanese Early Upper Paleolithic (EUP) technologies appeared in different regions of the archipelago around 38,000 years ago accompanied by remains of Homo sapiens found in Okinawa Island (c.36,000 cal BP) [@yamaoka2012use; @izuho201421]. These technologies include trap pits, cobble concentrations, hearths, charcoal concentrations, and toolkits including trapezoids, pointed-shaped backed blades, backed points, burins, end scrapers, side scrapers, wedges, beak-shaped tools, axes, edge-ground axes, hammerstones, cobble tools, and anvils [@bae2017origin; @izuho201421]. These innovations have been interpreted as evidence of new foraging methods such as watercraft (marine transport of obsidian from the Kozu Island) and bow-and-arrow technology, driven by increased population and ecological changes [@yamaoka2012use; @bae2017origin; @nakazawa2018quaternary; @morisaki2019early]. Hafted trapezoids were likely multifunctional tools adapted to the specific environmental settings of the different Japanese Islands [@ono2002radiocaron; @yamaoka2012use]. Ishinomoto 8-ku is one of the earliest sites dated to 39,690--34,790 cal BP, located in Takuma upland of Kumamoto Prefecture. A total of 500 stone artefacts were discovered including trapezoids, side scrapers, flakes, flake cores, edge-ground axes, and cobble tools. The main lithic raw materials are chert and andesite [@izuho201421]. Stemmed points appear much later, dated to between about 15,500 and 13,800 cal BP [@ono2002radiocaron; @tsutsumi2007dynamics].

What makes the Korean Late Pleistocene technological transition distinctive from what we see at Shuidonggou and in Japan is that the earliest signs of new technologies in Korean assemblages are stemmed points, followed by blade technologies. Stemmed points appeared in both regions after the case of Korean Peninsula. The oldest stemmed points so far are from the Yonghodong site dated to 38.5 ka and made on elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin]. After blades appeared at c. 27 ka and were used to make the stemmed points, the shape of points became more standardized with one or two ridges on the dorsal side and triangular cross section. Replacing flakes with blades as blanks of the stemmed points led to an increase in quantity of the points over time. In addition to the stemmed points, flake tools became more complex, with a greater diversity in both type and size along with continuously used core tools during the Late Pleistocene [@bae2010origin; @lee2013current; @Lee_2016]. There are several attempts to explain this technological transition and they can be summarized into two competing models: 'heterogenic' migration [@bae2010origin] or *in situ* evolution [@seong2009emergence]. The migration model argues that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins as the result of continuing influx of modern human migration from two routes. Bae (2010) explains that blade and stemmed points arrived in the Korean peninsula from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP with the earliest evidence of blades in those regions. He assumes that core and flake industry come from southern China based on the existence of similar assemblages in both Korea and southern China. In addition, the genetic studies of two “Asian-specific” Y-chromosome haplogroups (O3-M122, D-M174), mtDNA and autosomal SNPs, and other laboratories verify that modern humans initially arrive in southern China and the population migrated to northward [@bae2013early; @lee2013current]. The '*in situ*' model claims that stemmed points and other Late Paleolithic assemblages including blade industries autonomously emerged in the south of the Korean peninsula with no external influence [@seong2008tanged]. This claim is supported by Korea having the earliest appearance of stemmed points in Northeast Asia, at sites such as Hwadaeri, Hopyeongdong, and Yonghodong dated to 40-35ka BP [@seong2009emergence]. Seong(2009) uses the increased blade-to-flake ratios on lithic assemblages to support the *in situ* model based on the premise that the blade industry represents new technology while flakes indicate a continuing local one.

@@

Our brief review of Late Pleistocene technological innovations in Northeast Asia shows that previous work has largely focussed on timing, location, origins and description of these new tools, but modelling the land-use behaviours associated with these assemblages has rarely been undertaken [@seong2008tanged; @seong2009emergence; @bae2010origin; @lee2013current; @Lee_2016]. We focus on the appearance of stemmed points in Korea because these are a new technology that represents a new hunting method. This is because the hafting implies throwing techniques that can reach long-distance targets [@Keeley_1982; @kuhn2015artifacts]. We draw on behavioural ecological theory to model the effects of this technological change. Behavioural ecology theory offers structured models and testable hypotheses based on optimality assumptions that stand on economic rationality and environmental knowledge [@winterhalder1992evolutionary; @prentiss2019handbook]. The patch choice model built around the marginal value theorem predicts that forager mobility between patches and inter-patch travel times will increase when resource patch productivity decreases [@smith1991inujjuamiunt; @smith1983anthropological; @bettinger1997evolutionary; @wolverton2015patch; @llano2015optimal]. We can model individual artefact types as a kind of resource patch. Foragers may find a more costly and complicated technology (such as stemmed points relative to flakes) optimal if they have a long cumulative time in use, if they are maintainable and reusable in a landscape where lithic raw material supply is uncertain [@torrence1983time; @bleed1986optimal]. That is, they may move to a patch where they can stay longer (i.e. an artefact type that has a long and extendible use-life) when they are not certain of the productivity and travel times of other patches [@kuhn2015artifacts].

We ask two questions to investigate the period around the introduction of stemmed points in Korea. First, what changes in foragers' landscape use were associated with the introduction of the new tools? Specifically, was the use of stemmed points associated with occupation of marginal or extreme environments? Our assumption here is that this new technology might have allowed foragers to explore and sojourn in less productive landscapes by increasing their mobility. In Australia during the mid-Holocene, Hiscock et al. argue that the first appearance of microlithics in many locations represent a portable and multi-functional toolkit that minimize travel expenses and increase tool readiness to adapt to patchy and unpredictable resources associated with environmental change [@Hiscock_1994; @Hiscock_Clarkson_Mackay_2011].Similarly, in Japan, Morisaki et al. [@Morisaki_Izuho_Terry_Sato_2015] argue that the transition from trapezoid to the blades and projectile points around 25 ka cal BP in north Paleo-Honshu Island enabled the foragers to extend their occupation into cold grassland landscapes.

Second, we ask what changes in the way people use habitation sites and mobility were associated with the new technology? After the appearance of these new tools, did people tend to be more or less mobile? The stone artefact assemblage from a site can inform us how long individuals or groups stayed and give insights into their activities [@binford1979organization; @kelly1995foraging; @kuhn2016moves; @holdaway2019surface]. For example, stone artefact density and the frequency of retouched pieces (scaled to the volume of excavated sediment) can be used as proxies to represent occupation patterns [@barton2011modeling; @clark2017lithics]. Assemblages with low density but a high proportion of retouched and backed pieces may indicate the remains of 'short-term camp', which is a site for small and ephemeral overnight camp or limited activity station. On the other hand, the combination of high density and low retouched pieces might be expected at logistically organized basecamp, which is a site with greater residential stability, long site occupation and occupied by larger groups [@clark2017lithics; @clark2019landscapes].

@@

After the first excavation of a Paleolithic site in Korea at Seokjang-ri site in the 1960s, more than 200 Paleolithic sites have been discovered in South Korea [@lee2019were]. We selected sites dated to before and after the transition period (40-35 ka) to analyse the process of technological change. This resulted in a sample of 33 assemblages from 22 sites spanning 49-24 ka (Figure \@ref(fig:korea-map)). We identified multiple assemblages in a site where culturally sterile deposits separated artefact-bearing deposits, or where stratigraphic units could be identified by major differences in the texture and composition of the sedimentary deposit containing the stone artefacts. For example, the Hwadaeri site has three cultural layers. The lowest horizon at the bottom, dated to 39,000±1400 BP by OSL, contains coarse flake tools made of vein quartz and quartzite. The middle cultural horizon has stemmed points made of porphyry, with quartz and quartzite dominating the assemblage. This layer was dated to 31,200±900 BP by radiocarbon dating and 30,000±1,700 BP by OSL. The uppermost layer, dated to 22,000±100 BP by OSL, contains blades, scrapers, awls, and denticulates [@seong2009emergence]. We obtained the data from published excavation reports held by provincial museums.

@@

The 33 assemblages consist of (non-blade) cores, blade cores, blades, flakes, debris, hammers, choppers, planes, polyhedrals, side scrapers, end scrapers, notches, cleavers, stemmed points, awls, denticulates, burins, handaxes, blanks, knives, handadzes, peaks, flatters, flakers etc. We excluded some artefacts from our sample because of uncertainty about their typology. The stone artefacts from all of these sites are stored in museums located throughout South Korea, and have been briefly described in the excavation reports. We have visited most of these collections to obtain permission to study the materials.

@@

The artefact volumetric density is defined as the total number of artefacts per cubic meter of excavated sediment, and serves as a proxy for the accumulation ratio of artefacts [@clark2017lithics; @clark2019landscapes]. We used these density values to evaluate mobility and land-use practices of hunter-gatherers during the Late Pleistocene. The retouch frequency is the number of retouched tools divided by the total number of artefacts in an assemblage. This value represents mobility and duration of site occupation based on the assumption that tool reduction is a tactic to extend the use-life of a tool by generating sharp, usable edges while minimizing the cost of transporting tools or bulky raw materials by reducing the total load weight [@Kuhn_1990; @buvit2014late; @clark2019landscapes]. In other words, producing and carrying highly retouched tools will increase portability and efficiency by decreasing the size of both individual tools and the assemblages so that hunter-gatherers could move often or further, or both [@kuhn1991unpacking; @kuhn1994formal; @kuhn2004upper; @andrefsky1994raw; @davies2018modeling]. Combining these two proxies, the artefact volumetric density and the retouch frequency, we infer whether or not the hunter-gatherers were residentially mobile or logistically organized. If the assemblage in a site has a high artefact density and small proportion of retouch, we interpret it as a more expedient assemblage that represents "base camps" or "residences", while less density with higher proportions of retouch shows a more curated assemblage and indicates either residential mobility or certain task groups away from residential sites [@riel2004late; @clark2017lithics].

@@

Our study of raw material consumption patterns assumes that tool stone was transported and consumed to optimize mobility and minimize risk of not having artefacts when they were needed [@andrefsky1994raw; @brantingham2000raw; @brantingham2003neutral; @yue2020lithic]. The representation of raw materials in an assemblage can be an indicator of site occupation patterns. For example, a large variability of raw materials in a site can suggest long-term settlement, while varied and exotic raw materials can indicate higher mobility [@valde2020models]. Here we want to know the specific raw materials that relate to the appearance of a new technology, and understanding site occupation patterns by examining changes in the proportions of raw material types in the assemblages.

HBE models require environmental proxies to build hypotheses of behavioural change influenced by the surrounding environment. However, there are few paleoenvironmental proxies in Korea relevant to our target period. For example, geoarchaeological observations simply suggest that environment of the period was overall cooler and drier during MIS 3 [@Han_2008; @seong2008tanged; @Choi_2011; @chang2013human; @im2015study; @bak2017late]. To supplement the limited availability of high-resolution local proxies, we used climate information from the simulated data set of paleoclimate including global monthly temperature, covering the last 120,000 years [@beyer2020high]. We applied information of site location (all open-sites) including elevation, latitude and longitude to the simulated data set to get the annual temperature of our research period and area. Besides the climate analysis, we analysed site elevation as a proxy to represent different environments at local scales, to identify how these impact mobility and site occupation. We mainly focused on the elevation pattern of the sites before and after the appearance of new technology and its relationship with the stemmed points.

We used radiocarbon dates as a reference for population level. Summed probability distributions (SPD) of radiocarbon dates are increasingly used to infer temporal trends in past human populations [@rick1987dates; @bamforth2012radiocarbon; @shennan2013regional; @timpson2014reconstructing; @contreras2014summed; @riris2019widespread]. Despite limitations of the validity of SPD [@williams2012use; @bamforth2012radiocarbon], including the lack of excavated sites in certain periods, sample size, and calibration effects, this method may allow a first approximation of population fluctuations. Although this is often used with samples of thousands of ages, this method can also be useful for small sample sizes. For instance, Timpson et al. apply SPD to 93 ages from Eastern Middle Sweden and demonstrate that the result is equivalent to a larger sample size (n= 243) [@timpson2014reconstructing]. We used 108 radiocarbon ages from 22 sites to examine connections between population trends during the Late Pleistocene in Korea and technological innovation.

There is a debate about whether past population growth has played a role in technological change [@shennan2001demography; @kline2010population]. @shennan2009pattern observed a correlation between population decrease and technological simplification in his study of lithic arrowheads from the southern Scandinavian Mesolithic. He found that the lithics became less complex, and point shape changed as the population level dropped. On the other hand, @collard2013population argue that regional archaeological data do not support the population-driven models [@buchanan2016drivers]. For example, @buchanan2016drivers concluded that changes in point types from \>13,000 years ago to 400 years ago in Texas were related to environmental risk, and not dependent on population size. We explored SPD to test for correlations between population, environmental proxies and technological change. We used the R package 'rcarbon' [@cremaanalysing] to perform statistical analyses and conduct model testing using Monte Carlo methods [@timpson2014reconstructing; @crema2016summed; @palmisano2020holocene]. We generated exponential, uniform, linear, and logistic models and evaluated the model goodness-of-fit by applying Akaike's Information Criterion (AIC) as a selection criterion [@sakamoto1986akaike; @bevan2017holocene; @riris2019widespread].
